\chapter{Conclusions}

We extendedn traditional Hoare logic into a more abstract and versatile 
framework. By incorporating principles from order theory and abstract 
interpretation, we developed a robust method for reasoning about a broader range 
of program properties. This abstract Hoare logic framework maintains the 
foundational principles of soundness and completeness while expanding its 
applicability.

We have shown that multiple program logics known in literature are actually 
special cases of Abstract Hoare logic, while constructing a program logic for 
hyperproperties in the framework of Abstract Hoare logic we also provided a 
compositional definition of strongest hyper postcondition.


Additionally, we have introduced the concept of Reverse Abstract Hoare logic, 
which enhances the framework by enabling backward reasoning by reusing the ideas
from Abstract Hoare logic. 

Throughout this work, we have demonstrated the practical applications of our 
framework through various examples and instantiations. These examples illustrate 
the versatility and effectiveness of the Abstract Hoare logic framework.

\section{Future work}
The following are only preliminary results that we weren't able to explore
because of time constraints.

\subsection{Under-approximating logics}
Hoare-style logics have not only been used to perform over-approximate
reasoning. Incorrectness logic \cite{OHearn19} uses the same framework as Hoare
logic but focuses on under-approximation. An incorrectness triple is defined as
$\htriple{P}{C}{Q} \iff \sem{C}(P) \supseteq Q$. The definition of easy
represents the same in Abstract Hoare logic. Validity differs from Hoare logic
in the direction of set inclusion, so we could easily provide the definition
for Abstract Incorrectness Logic triples $\atriple{P}{C}{Q} \iff \asem{C}(P)
\geq_A Q$ and perform the same translation that we did for the rules of Hoare
Logic for the rules of Incorrectness logic.

\subsection{Hyper domains}
We used hyper domains to construct program logics that that are able to reason
about hyperproperties, but we didn't explore what kind of calculi proofsystem 
they create when they are used to instantiate Reverse Hoare logic and on the 
opposite lattice.

\section{Related work}

The idea of systematically constructing program logics is not new. Kleene
Algebra with Tests (KAT) \cite{Kozen97} was one of the first of this kind of work.
In section \ref{chp:join-meet-rules} we have discussed how in general we
cant distribute non-deterministic choice ($\asem{(C_1 + C_2) \fcmp C_3} \neq
\asem{(C_1 \fcmp C_3) + (C_2 \fcmp C_3)}$) violating one of the axioms of
Kleene algebras; Another similar alternative was proposed in \cite{Martin06}
using traced monoidal categories to encode properties of the program, for
example the monoidal structure is used to model the non-deterministic choice,
but imposes the same distributivity requirements of Kleene Algebras (this is
caused by $\oplus$ being a bifunctor), but disregarding expressivity, the
main difference it's in the philosophy behind the approach, Abstract Hoare logic
is a more semantics centered approach instead of being an "equational" theory
like KAT. The more semantics centered approach was also vital in seeing
providing the idea that the abstract inductive semantics could be used not only
to encode the strongest postcondition but also the strongest liberal postcondition,
weakest precondition and weakest liberal precondition and unify all the
partial hoare-like logics.

A more similar approach to rhe one of Abstract Hoare Logic is Outcome Logic
\cite{Zilberstein23} as like Abstract Hoare Logic the semantics of the language
is parametric on the domain of execution but the assertion language is fixed
if we ignore the basic assertions on program states, Outcome logic had as original
goal the of of unifiyng correctness and incorrectness reasoning with the powerset
instantiation not to be a minimal theory for sound and complete Hoare-like
logics, in fact the relative completeness proof is missing, as already discussed
in \cite{Darnier2023} Outcome logic with the powerset instantiation is actually
a proofsystem for 2-hyperproperties (hyperproperties regarding at most two
executions) thus Outcome triples can be proved in the instantiation of
Abstract Hoare Logic provided in section \ref{chp:hyper} even tough it would
be interesting to find a direct encoding of Outcome Logic in terms Abstract
Hoare Logic.
