\chapter*{Introduction}

The verification of program correctness is a critical and crucial task in
computer science. Ensuring that software behaves as expected under all possible
conditions is fundamental in a society that increasingly relies on computer
programs. Software engineers often reason about the behavior of their programs
at an intuitive level. While this is definitely better than not reasoning at
all, intuition alone becomes insufficient as the size of programs grows.

Writing tests for programs is definitely a useful task, but at best, it can
show the presence of bugs, not prove their absence. We cannot feasibly write
tests for every possible input of the program. To offer a guarantee of the
absence of undesired behaviors, we need sound logical models rooted in logic.
The field of formal methods in computer science aims at developing the logical
tools necessary to prove properties of software systems.

Hoare logic, first introduced by Hoare in the late 60s \cite{Hoare69}, provides
a set of logical rules to reason about the correctness of computer programs.
Hoare logic formalizes, with axioms and inference rules, the relationship
between the initial and final states after executing a program.

Hoare logic, beyond being one of the first program logics, is arguably also one
of the most influential ideas in the field of software verification. It created
the whole field of program logics—systems of logical rules aimed at proving
properties of programs. Over the years, modifications of Hoare logic have been
developed, sometimes to support new language features such as dynamic memory
allocation and pointers, or to prove different properties such as equivalence
between programs or properties of multiple executions. Every time Hoare logic
is modified, it is necessary to prove again that the proof system indeed proves
properties about the program (soundness) and ideally that the proof system is
powerful enough to prove all the properties of interest (completeness).

Most modifications of Hoare logic usually do not alter the fundamental proof
principles of the system. Instead, they often extend the assertion language to
express new properties and add new commands to support new features in
different programming languages.

In this work, we introduce Abstract Hoare Logic, which aims to be a framework
general enough to serve as an extensible platform for constructing new
Hoare-like logics without the burden of proving soundness and completeness
anew. We demonstrate, through examples, how some properties that are not
expressible in standard Hoare logic can be simply instantiated within Abstract
Hoare Logic, while keeping the proof system as simple as possible.

The theory of Abstract Hoare Logic is deeply connected to the theory of
abstract interpretation \cite{Cousot77}. The semantics of the language is
defined as an inductive abstract interpreter, and the validity of the Abstract
Hoare triples depends on it. Since we do not use the strongest postcondition
directly, we are able to reason about properties that are not expressible in
the powerset of the program states, such as hyperproperties.

This thesis is structured as follows:
\begin{itemize}
  \item In Chapter 1, we introduce the basic mathematical background of order
    theory and abstract interpretation.

  \item In Chapter 2, we introduce standard Hoare logic and the general
    framework of Abstract Hoare Logic: the extensible language $\lang$, its
    syntax and semantics, the generalization of the strongest postcondition,
    and finally, Abstract Hoare Logic and its proof system, proving the general
    results of soundness and relative completeness.

  \item In Chapter 3, we show some notable instantiations of Abstract Hoare
    Logic: we demonstrate that it is possible to obtain program logics where
    the implication is decidable, thus making the goal of checking a derivation
    computable; we show how to obtain a proof system for hyperproperties (and
    we introduce the concept of the strongest hyper postcondition); Finally, we
    show that it is possible to obtain a proof system for partial
    incorrectness.

  \item In Chapter 4, we show how to enrich the barebones proof system of
    Abstract Hoare Logic by adding more restrictions on the assertion language
    or the semantics.

  \item In Chapter 5, we show how to reuse the idea of Abstract Hoare Logic to
    generalize proof systems for backward reasoning.

  \item In Chapter 6, we provide a brief summary of the most important
    contributions of the thesis. We discuss possible extensions to the
    framework of Abstract Hoare Logic and, to conclude, we examine the
    relationship of Abstract Hoare Logic with other similar work.
\end{itemize}

\chapter{Background}

\subimport{./}{order_theory}
\subimport{./}{abstract_interpretation}
